---
title: "tree_orderer2"
author: "hirgh815"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Fist I want to clarify the difference between these two function I found:

## ape rotateConstr() and phytools minRotate()

The rotateConstr() function rotates internal nodes to find the closest possible match to the desired_order vector without altering the underlying tree topology (i.e., it doesn't change who is related to whom, just how it's visualized).
The minRotate() is from phytools. It also leaves the topology unchanged.
Instead of directly following the supplied ordering, it searches over allowable node rotations to minimize disagreement between the current tip order and the desired tip order.

The disagreement is measured by the number of pairwise inversions. For small tress they are equal, but these algorithm differ when we encounter cases where more than one way exists to rotate a tree without changing its evolutionary relationships, and the software must choose one arrangement.

```{r}
#install.packages("ape")
library(ape)
library(phytools)

# Create a random tree with 5 tips
set.seed(42)
my_tree <- rtree(5, tip.label = c("A", "B", "C", "D", "E"))

# Let's see the current tip order
my_tree$tip.label 
# Output is: "A" "B" "C" "D" "E" (but visual order depends on topology)

# Define your desired target order
desired_order <- c("E", "C", "A", "D", "B")

# Reorder the tree topology to match the constraint
reordered_tree_1 <- rotateConstr(my_tree, desired_order)
reordered_tree_2 <- minRotate(my_tree, desired_order)

# Check the new visual tip order
#reordered_tree$tip.label

par(mfrow = c(1, 3)) # Side-by-side plotting

# Original Tree
plot(my_tree, main = "Original Tree")

# Reordered Tree using ape
plot(reordered_tree_1, main = "Match to Desired Order 1")

# Reordered Tree using phytools
plot(reordered_tree_2, main = "Match to Desired Order 2")

print(all.equal(reordered_tree_1,reordered_tree_2))

```

## Solution 1

Originally I gave a complete ordering (desired_order) that was already known. Now, the ordering itself has to be inferred from the trait values, while respecting the topology of the tree. The PCA algorithm can be a heuristic for generating such an ordering but it is also very approximate.

```{r}
library(ape)
library(phytools)

traits <- matrix(runif(30, 0, 5), nrow = 5, dimnames = list(my_tree$tip.label))

pc <- prcomp(traits, scale.=TRUE)

score <- pc$x[,1]

desired_order <- names(sort(score))

desired_order


tree_Rotate <- rotateConstr(my_tree, desired_order) # or minRotate(my_tree, desired_order)

par(mfrow=c(1,2))

plot(my_tree,
     main="Original")

plot(tree_Rotate,
     main="Rotate (PC1 order)")
```

## Solution 2

Another robust mathematical approach is seriation. This treats the problem of ordering the tips as a Traveling Salesperson Problem (TSP), find the shortest path (1D order again) that visits every single tip's trait vector exactly once, minimizing the step-by-step distance.

```{r}
# install.packages("seriation")
library(ape)
library(seriation)

# Simulate a 3D trait matrix
dist_matrix <- dist(traits)

# Use TSP solver to find the optimal 1D path minimizing total step-wise distance
tsp_sol <- seriate(dist_matrix, method = "TSP")
suggested_order <- get_order(tsp_sol) # Extract the 1D permutation

# Map the indices back to tip labels
target_sequence <- labels(dist_matrix)[suggested_order]

# Rotate the tree to match this optimal TSP sequence
optimized_tree <- rotateConstr(my_tree, target_sequence) # or minRotate(my_tree, target_sequence)

par(mfrow=c(1,2))

plot(my_tree,
     main="Original")

plot(optimized_tree,
     main="Rotate (TSP order)")
```


## Solution 3

For such a small tree I can explore everything:

```{r}

## Cost function

# ordering_cost <- function(order, traits){
# 
#     X <- traits[order,,drop=FALSE]
# 
#     sum(sqrt(rowSums((X[-1,]-X[-nrow(X),])^2)))
# 
# }

# 1. Get plot order safely via .PlotPhyloEnv
get_plot_order <- function(tree){
    oldpar <- par(no.readonly = TRUE)
    on.exit(par(oldpar))

    plot(tree, plot = FALSE)
    pp <- get("last_plot.phylo", envir = .PlotPhyloEnv)
    n <- length(tree$tip.label)
    
    # Order tips based on their vertical positioning (yy) in the plot
    tree$tip.label[order(pp$yy[1:n])]
}

# 2. Calculate continuous trait transition distance along the tip profile
trait_cost <- function(tree, traits){
    ord <- get_plot_order(tree)
    X <- traits[ord, , drop = FALSE]
    
    sum(
        sqrt(
            rowSums((X[-1, , drop = FALSE] - X[-nrow(X), , drop = FALSE])^2)
        )
    )
}

# 3. Generate every valid topological rotation using a Breadth-First Search (BFS)
all_rotations <- function(tree){
    Ntip <- length(tree$tip.label)
    internal <- (Ntip + 1):(Ntip + tree$Nnode)
    
    # Initialize queue and visited set using Newick format strings
    initial_key <- write.tree(tree)
    seen <- initial_key
    queue <- list(tree)
    all_trees <- list(tree)
    
    head_idx <- 1
    while(head_idx <= length(queue)){
        current_tree <- queue[[head_idx]]
        head_idx <- head_idx + 1
        
        for(node in internal){
            # Use ape::rotate to flip the chosen internal node
            tmp <- rotate(current_tree, node)
            key <- write.tree(tmp)
            
            if(!(key %in% seen)){
                seen <- c(seen, key)
                queue[[length(queue) + 1]] <- tmp
                all_trees[[length(all_trees) + 1]] <- tmp
            }
        }
    }
    return(all_trees)
}

# 4. Evaluate every single unique rotation configuration
trees <- all_rotations(my_tree)
costs <- numeric(length(trees))
orders <- vector("list", length(trees))

for(i in seq_along(trees)){
    # FIX: Pass the tree object to trait_cost, not the raw text vector of labels
    costs[i] <- trait_cost(trees[[i]], traits)
    orders[[i]] <- get_plot_order(trees[[i]])
}

# Find the global minimum configuration
best <- which.min(costs)
best_tree <- trees[[best]]
best_order <- orders[[best]]
best_cost <- costs[[best]]

# Output results
cat("Best order:\n")
print(best_order)
cat("\nCost =", best_cost, "\n")

# Layout comparison plots
par(mfrow = c(1, 2))
plot(my_tree, main = paste("Original (Cost:", round(trait_cost(my_tree, traits), 2), ")"))
plot(best_tree, main = paste("Optimized (Cost:", round(best_cost, 2), ")"))
```

## solution 4

For larger trees (e.g., 50–1000 tips), the brute-force enumeration above becomes infeasible. So instead I use meta-heuristic. Practical approach is to Start from the current tree embedding.
Randomly rotate internal nodes (swap child branches).
Accept better solutions always; worse solutions with probability $p=exp(-\frac{\Delta}{T})$
where T decreases over time (simulated annealing). Also Stop early if no improvement occurs for many iterations, temperature becomes very small, or maximum iterations reached.

```{r}
set.seed(42)

my_tree <- rtree(
    100,
    tip.label=paste0("Tip",1:100)
)

traits <- matrix(
    runif(100*6,0,5),
    nrow=100,
    dimnames=list(my_tree$tip.label,
                  paste0("Trait",1:6))
)



# Cost function


# trait_cost <- function(tree, traits){
# 
#     order <- tree$tip.label
# 
#     X <- traits[order,,drop=FALSE]
# 
#     sum(
#         sqrt(
#             rowSums(
#                 (X[-1,]-X[-nrow(X),])^2
#             )
#         )
#     )
# }

# Fast tree tip order (no plotting)


get_plot_order <- function(tree){

    n <- length(tree$tip.label)

    # children of each internal node
    children <- split(
        tree$edge[,2],
        tree$edge[,1]
    )

    root <- setdiff(
        tree$edge[,1],
        tree$edge[,2]
    )[1]


    # recursive depth-first traversal
    traverse <- function(node){

        # tip
        if(node <= n){
            return(node)
        }

        unlist(
            lapply(
                children[[as.character(node)]],
                traverse
            )
        )
    }


    idx <- traverse(root)

    tree$tip.label[idx]
}


trait_cost <- function(tree, traits){

    ord <- get_plot_order(tree)

    X <- traits[ord,,drop=FALSE]

    sum(
        sqrt(
            rowSums(
                (X[-1,]-X[-nrow(X),])^2
            )
        )
    )

}


# Random rotation of one internal node


random_rotate <- function(tree){

    Ntip <- length(tree$tip.label)

    internal_nodes <- (Ntip+1):(Ntip+tree$Nnode)

    node <- sample(internal_nodes,1)

    tree2 <- tree

    children <- which(tree2$edge[,1]==node)

    if(length(children)==2){

        tree2$edge[children,2] <-
            rev(tree2$edge[children,2])

    }

    tree2

}




# Simulated annealing optimizer


anneal_tree_order <- function(
        tree,
        traits,
        max_iter=100000,
        temperature=1,
        cooling=0.99995,
        patience=5000
){

    current_tree <- tree

    current_cost <- trait_cost(
        current_tree,
        traits
    )

    cat(
    "Starting cost:",
    current_cost,
    "\n"
    )

    best_tree <- current_tree

    best_cost <- current_cost


    no_improve <- 0


    for(iter in 1:max_iter){


        candidate <- random_rotate(
            current_tree
        )


        candidate_cost <- trait_cost(
            candidate,
            traits
        )


        delta <- candidate_cost-current_cost



        # Metropolis acceptance rule


        accept <- FALSE

        if(delta < 0){

            accept <- TRUE

        } else {

            probability <- exp(
                -delta/temperature
            )

            if(runif(1)<probability){
                accept <- TRUE
            }

        }


        if(accept){

            current_tree <- candidate

            current_cost <- candidate_cost

        }



        # Save best


        if(current_cost < best_cost){

            best_tree <- current_tree

            best_cost <- current_cost

            no_improve <- 0

        } else {

            no_improve <- no_improve + 1

        }



        # Cooling


        temperature <- temperature * cooling




        # Early stopping


        if(no_improve >= patience){

            message(
                "Stopped early at iteration ",
                iter
            )

            break

        }


        if(temperature < 1e-8){

            message(
                "Temperature exhausted"
            )

            break

        }


        if(iter %% 1000 == 0){

            message(
                "Iteration ",
                iter,
                " Best cost = ",
                best_cost
            )

        }

    }


    list(
        tree=best_tree,
        cost=best_cost,
        iterations=iter
    )

}




# Run optimization


result <- anneal_tree_order(
    my_tree,
    traits,
    max_iter=200000,
    temperature=10,
    cooling=0.99997,
    patience=10000
)



# Results


optimized_tree <- result$tree

init_cost <- trait_cost(my_tree,traits)
cat(
    "Initial cost:",
    init_cost,
    "\n"
)


cat(
    "Final cost:",
    result$cost,
    "\n"
)

cat(
    "Iterations:",
    result$iterations,
    "\n"
)



# Plot example


par(mfrow=c(1,2))

plot(my_tree,
     show.tip.label=FALSE,
     main="Original")

plot(optimized_tree,
     show.tip.label=FALSE,
     main="Optimized")
```


## Solution 5

In the last section, I compare simulated annealing for two strategies. Random initialization vs informed initialization. Instead of starting from an arbitrary tree embedding, I start near a (biologically?) meaningful arrangement, but random start had a better result at the cost of many more iterations. Simulated annealing then only needs to make local corrections. Experiment:
Random initialization + annealing
PCA + rotateConstr initialization + annealing
```{r}



# Simulate tree and traits


set.seed(42)

nTips <- 500

my_tree <- rtree(
    nTips,
    tip.label=paste0("Tip",1:nTips)
)


traits <- matrix(
    runif(nTips*10,0,5),
    nrow=nTips,
    dimnames=list(
        my_tree$tip.label,
        paste0("Trait",1:10)
    )
)



############################################################
# Objective function


# Fast tree tip order (no plotting)


get_plot_order <- function(tree){

    n <- length(tree$tip.label)

    # children of each internal node
    children <- split(
        tree$edge[,2],
        tree$edge[,1]
    )

    root <- setdiff(
        tree$edge[,1],
        tree$edge[,2]
    )[1]


    # recursive depth-first traversal
    traverse <- function(node){

        # tip
        if(node <= n){
            return(node)
        }

        unlist(
            lapply(
                children[[as.character(node)]],
                traverse
            )
        )
    }


    idx <- traverse(root)

    tree$tip.label[idx]
}


trait_cost <- function(tree, traits){

    ord <- get_plot_order(tree)

    X <- traits[ord,,drop=FALSE]

    sum(
        sqrt(
            rowSums(
                (X[-1,]-X[-nrow(X),])^2
            )
        )
    )

}




# Random rotation


random_rotate <- function(tree){

    Ntip <- length(tree$tip.label)

    node <- sample(
        (Ntip+1):(Ntip+tree$Nnode),
        1
    )

    rows <- which(tree$edge[,1]==node)

    if(length(rows)==2){

        tree$edge[rows,2] <-
            rev(tree$edge[rows,2])

    }

    tree
}




# Annealing


anneal_tree <- function(
        tree,
        traits,
        max_iter=100000,
        temperature=10,
        cooling=.99995,
        patience=5000
){


    current <- tree

    current_cost <- trait_cost(
        current,
        traits
    )
    
    cat(
    "Starting cost:",
    current_cost,
    "\n"
    )

    best <- current
    best_cost <- current_cost


    no_change <- 0


    for(i in 1:max_iter){


        candidate <- random_rotate(current)


        candidate_cost <- trait_cost(
            candidate,
            traits
        )


        delta <- candidate_cost-current_cost


        if(
            delta < 0 ||
            runif(1)<exp(-delta/temperature)
        ){

            current <- candidate
            current_cost <- candidate_cost

        }


        if(current_cost < best_cost){

            best <- current
            best_cost <- current_cost

            no_change <- 0

        } else {

            no_change <- no_change+1

        }


        temperature <- temperature*cooling


        if(no_change > patience){

            return(
                list(
                    tree=best,
                    cost=best_cost,
                    iterations=i
                )
            )

        }

    }


    list(
        tree=best,
        cost=best_cost,
        iterations=max_iter
    )

}




############################################################
# METHOD 1:
# Random starting layout
############################################################

cat(
    "Initial cost:",
    trait_cost(my_tree,traits),
    "\n"
)

cat("\nRandom initialization\n")

random_order <- sample(my_tree$tip.label)

random_tree <- rotateConstr(
    my_tree,
    random_order
)

cat(
    "Initial random cost:",
    trait_cost(random_tree, traits),
    "\n"
)

start1 <- Sys.time()



random_result <- anneal_tree(
    random_tree,
    traits
)

time1 <- Sys.time()-start1



############################################################
# METHOD 2:
# PCA initialization
############################################################

cat("\nPCA initialization\n")


############################################################
# PCA scores
############################################################

pc <- prcomp(
    traits,
    scale.=TRUE
)


pc_order <- rownames(
    traits
)[
    order(pc$x[,1])
]


############################################################
# Rotate tree to PCA order
############################################################

pca_tree <- rotateConstr(
    my_tree,
    pc_order
)





cat(
    "Initial PCA cost:",
    trait_cost(pca_tree,traits),
    "\n"
)



start2 <- Sys.time()

pca_result <- anneal_tree(
    pca_tree,
    traits
)

time2 <- Sys.time()-start2



############################################################
# Compare
############################################################

results <- data.frame(

    Method=c(
        "Random start",
        "PCA + rotateConstr"
        ),

    Initial_cost=c(
        trait_cost(random_tree, traits),
        trait_cost(pca_tree, traits)
        ),

    Final_cost=c(
        random_result$cost,
        pca_result$cost
    ),

    Iterations=c(
        random_result$iterations,
        pca_result$iterations
    ),

    Time_seconds=c(
        as.numeric(time1),
        as.numeric(time2)
    )

)

print(results)
```

