---
title: "tree_orderer2"
author: "hirgh815"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Fist I want to clarify the difference between these two function I found:

## ape rotateConstr() and phytools minRotate()

The rotateConstr() function rotates internal nodes to find the closest possible match to the desired_order vector without altering the underlying tree topology (i.e., it doesn't change who is related to whom, just how it's visualized).
The minRotate() is from phytools. It also leaves the topology unchanged.
Instead of directly following the supplied ordering, it searches over allowable node rotations to minimize disagreement between the current tip order and the desired tip order.

The disagreement is measured by the number of pairwise inversions. For small tress they are equal, but these algorithm differ when we encounter cases where more than one way exists to rotate a tree without changing its evolutionary relationships, and the software must choose one arrangement.

```{r}
#install.packages("ape")
library(ape)
library(phytools)

# Create a random tree with 5 tips
set.seed(42)
my_tree <- rtree(5, tip.label = c("A", "B", "C", "D", "E"))

# Let's see the current tip order
my_tree$tip.label 
# Output is: "A" "B" "C" "D" "E" (but visual order depends on topology)

# Define your desired target order
desired_order <- c("E", "C", "A", "D", "B")

# Reorder the tree topology to match the constraint
reordered_tree_1 <- rotateConstr(my_tree, desired_order)
reordered_tree_2 <- minRotate(my_tree, desired_order)

# Check the new visual tip order
#reordered_tree$tip.label

par(mfrow = c(1, 3)) # Side-by-side plotting

# Original Tree
plot(my_tree, main = "Original Tree")

# Reordered Tree using ape
plot(reordered_tree_1, main = "Match to Desired Order 1")

# Reordered Tree using phytools
plot(reordered_tree_2, main = "Match to Desired Order 2")

print(all.equal(reordered_tree_1,reordered_tree_2))

```

## Solution 1

Originally I gave a complete ordering (desired_order) that was already known. Now, the ordering itself has to be inferred from the trait values, while respecting the topology of the tree. The PCA algorithm can be a heuristic for generating such an ordering but it is also very approximate.

```{r}
library(ape)
library(phytools)

traits <- matrix(runif(30, 0, 5), nrow = 5, dimnames = list(my_tree$tip.label))

pc <- prcomp(traits, scale.=TRUE)

score <- pc$x[,1]

desired_order <- names(sort(score))

desired_order


tree_Rotate <- rotateConstr(my_tree, desired_order) # or minRotate(my_tree, desired_order)

par(mfrow=c(1,2))

plot(my_tree,
     main="Original")

plot(tree_Rotate,
     main="Rotate (PC1 order)")
```

## Solution 2

Another robust mathematical approach is seriation. This treats the problem of ordering the tips as a Traveling Salesperson Problem (TSP), find the shortest path (1D order again) that visits every single tip's trait vector exactly once, minimizing the step-by-step distance.

```{r}
# install.packages("seriation")
library(ape)
library(seriation)

# Simulate a 3D trait matrix
dist_matrix <- dist(traits)

# Use TSP solver to find the optimal 1D path minimizing total step-wise distance
tsp_sol <- seriate(dist_matrix, method = "TSP")
suggested_order <- get_order(tsp_sol) # Extract the 1D permutation

# Map the indices back to tip labels
target_sequence <- labels(dist_matrix)[suggested_order]

# Rotate the tree to match this optimal TSP sequence
optimized_tree <- rotateConstr(my_tree, target_sequence) # or minRotate(my_tree, target_sequence)

par(mfrow=c(1,2))

plot(my_tree,
     main="Original")

plot(optimized_tree,
     main="Rotate (TSP order)")
```


## Solution 3

For such a small tree I can explore everything:

```{r}

## Cost function

# ordering_cost <- function(order, traits){
# 
#     X <- traits[order,,drop=FALSE]
# 
#     sum(sqrt(rowSums((X[-1,]-X[-nrow(X),])^2)))
# 
# }

# 1. Get plot order safely via .PlotPhyloEnv
get_plot_order <- function(tree){
    oldpar <- par(no.readonly = TRUE)
    on.exit(par(oldpar))

    plot(tree, plot = FALSE)
    pp <- get("last_plot.phylo", envir = .PlotPhyloEnv)
    n <- length(tree$tip.label)
    
    # Order tips based on their vertical positioning (yy) in the plot
    tree$tip.label[order(pp$yy[1:n])]
}

# 2. Calculate continuous trait transition distance along the tip profile
trait_cost <- function(tree, traits){
    ord <- get_plot_order(tree)
    X <- traits[ord, , drop = FALSE]
    
    sum(
        sqrt(
            rowSums((X[-1, , drop = FALSE] - X[-nrow(X), , drop = FALSE])^2)
        )
    )
}

# 3. Generate every valid topological rotation using a Breadth-First Search (BFS)
all_rotations <- function(tree){
    Ntip <- length(tree$tip.label)
    internal <- (Ntip + 1):(Ntip + tree$Nnode)
    
    # Initialize queue and visited set using Newick format strings
    initial_key <- write.tree(tree)
    seen <- initial_key
    queue <- list(tree)
    all_trees <- list(tree)
    
    head_idx <- 1
    while(head_idx <= length(queue)){
        current_tree <- queue[[head_idx]]
        head_idx <- head_idx + 1
        
        for(node in internal){
            # Use ape::rotate to flip the chosen internal node
            tmp <- rotate(current_tree, node)
            key <- write.tree(tmp)
            
            if(!(key %in% seen)){
                seen <- c(seen, key)
                queue[[length(queue) + 1]] <- tmp
                all_trees[[length(all_trees) + 1]] <- tmp
            }
        }
    }
    return(all_trees)
}

# 4. Evaluate every single unique rotation configuration
trees <- all_rotations(my_tree)
costs <- numeric(length(trees))
orders <- vector("list", length(trees))

for(i in seq_along(trees)){
    # FIX: Pass the tree object to trait_cost, not the raw text vector of labels
    costs[i] <- trait_cost(trees[[i]], traits)
    orders[[i]] <- get_plot_order(trees[[i]])
}

# Find the global minimum configuration
best <- which.min(costs)
best_tree <- trees[[best]]
best_order <- orders[[best]]
best_cost <- costs[[best]]

# Output results
cat("Best order:\n")
print(best_order)
cat("\nCost =", best_cost, "\n")

# Layout comparison plots
par(mfrow = c(1, 2))
plot(my_tree, main = paste("Original (Cost:", round(trait_cost(my_tree, traits), 2), ")"))
plot(best_tree, main = paste("Optimized (Cost:", round(best_cost, 2), ")"))
```

## solution 4

For larger trees (e.g., 50–1000 tips), the brute-force enumeration above becomes infeasible. So instead I use meta-heuristic. Practical approach is to Start from the current tree embedding.
Randomly rotate internal nodes (swap child branches).
Accept better solutions always; worse solutions with probability $p=exp(-\frac{\Delta}{T})$
where T decreases over time (simulated annealing). Also Stop early if no improvement occurs for many iterations, temperature becomes very small, or maximum iterations reached.

```{r}
set.seed(42)

my_tree <- rtree(
    100,
    tip.label=paste0("Tip",1:100)
)

traits <- matrix(
    runif(100*6,0,5),
    nrow=100,
    dimnames=list(my_tree$tip.label,
                  paste0("Trait",1:6))
)



# Cost function


# get_plot_order <- function(tree){
# 
#     oldpar <- par(no.readonly = TRUE)
# 
#     on.exit(par(oldpar))
# 
#     plot(
#         tree,
#         plot=FALSE
#     )
# 
#     pp <- get(
#         "last_plot.phylo",
#         envir=.PlotPhyloEnv
#     )
# 
#     n <- length(tree$tip.label)
# 
#     tree$tip.label[
#         order(pp$yy[1:n])
#     ]
# }
# 
# trait_cost <- function(tree, traits){
# 
#     ord <- get_plot_order(tree)
# 
#     X <- traits[ord,,drop=FALSE]
# 
#     sum(
#         sqrt(
#             rowSums(
#                 (X[-1,]-X[-nrow(X),])^2
#             )
#         )
#     )
# }

# Fast tree tip order (Vectorized and fully compatible with ape structure)
get_plot_order <- function(tree) {
    n <- length(tree$tip.label)
    # Extracts tips in their exact plotting/traversal order instantly
    idx <- tree$edge[tree$edge[, 2] <= n, 2]
    tree$tip.label[idx]
}

trait_cost <- function(tree, traits) {
    ord <- get_plot_order(tree)
    X <- traits[ord, , drop = FALSE]
    
    sum(
        sqrt(
            rowSums(
                (X[-1, , drop = FALSE] - X[-nrow(X), , drop = FALSE])^2
            )
        )
    )
}

# Random rotation using safe and native ape infrastructure
random_rotate <- function(tree) {
    Ntip <- length(tree$tip.label)
    
    # Safe sampling of an internal node
    node <- sample((Ntip + 1):(Ntip + tree$Nnode), 1)
    
    # Properly rotates the node, preserving branch lengths and cladewise order
    ape::rotate(tree, node)
}

# Safely rebuilds the tree structure from scratch while preserving the visual layout
sanitize_tree <- function(tree) {
    ape::read.tree(text = ape::write.tree(tree))
}

# Annealing
anneal_tree <- function(
    tree,
    traits,
    max_iter = 100000,
    temperature = 10,
    cooling = .099995, # Adjusted to match step counts smoothly
    patience = 5000
) {
    current <- tree
    current_cost <- trait_cost(current, traits)
    
    cat("Starting cost:", current_cost, "\n")
    
    best <- current
    best_cost <- current_cost
    no_change <- 0
    
    for (i in 1:max_iter) {
        candidate <- random_rotate(current)
        candidate_cost <- trait_cost(candidate, traits)
        delta <- candidate_cost - current_cost
        
        if (delta < 0 || runif(1) < exp(-delta / temperature)) {
            current <- candidate
            current_cost <- candidate_cost
        }
        
        if (current_cost < best_cost) {
            best <- current
            best_cost <- current_cost
            no_change <- 0
        } else {
            no_change <- no_change + 1
        }
        
        temperature <- temperature * cooling
        
        if (no_change > patience) {
            return(
                list(
                    tree = best,
                    cost = best_cost,
                    iterations = i
                )
            )
        }
    }
    
    list(
        tree = best,
        cost = best_cost,
        iterations = max_iter
    )
}




# Run optimization
if (exists("anneal_tree_order")) rm(anneal_tree_order)

result <- anneal_tree(
    my_tree,
    traits,
    max_iter=200000,
    
    temperature=10,
    patience=10000
)



# Results


optimized_tree <- result$tree

init_cost <- trait_cost(my_tree,traits)
cat(
    "Initial cost:",
    init_cost,
    "\n"
)


cat(
    "Final cost:",
    result$cost,
    "\n"
)

cat(
    "Iterations:",
    result$iterations,
    "\n"
)



# Plot example


par(mfrow=c(1,2))

plot(my_tree,
     show.tip.label=FALSE,
     main="Original")

plot(optimized_tree,
     show.tip.label=FALSE,
     main="Optimized")
```


## Solution 5

In the last section, I compare simulated annealing for two strategies. Random initialization vs informed initialization. Instead of starting from an arbitrary tree embedding, I start near a (biologically?) meaningful arrangement, but random start had a better result at the cost of many more iterations. Simulated annealing then only needs to make local corrections. Experiment:
Random initialization + annealing
PCA + rotateConstr initialization + annealing
```{r}



# Simulate tree and traits


set.seed(42)

nTips <- 500

my_tree <- rtree(
    nTips,
    tip.label=paste0("Tip",1:nTips)
)


traits <- matrix(
    runif(nTips*10,0,5),
    nrow=nTips,
    dimnames=list(
        my_tree$tip.label,
        paste0("Trait",1:10)
    )
)



############################################################
# Objective function


# Fast tree tip order (Vectorized and fully compatible with ape structure)
get_plot_order <- function(tree) {
    n <- length(tree$tip.label)
    # Extracts tips in their exact plotting/traversal order instantly
    idx <- tree$edge[tree$edge[, 2] <= n, 2]
    tree$tip.label[idx]
}

trait_cost <- function(tree, traits) {
    ord <- get_plot_order(tree)
    X <- traits[ord, , drop = FALSE]
    
    sum(
        sqrt(
            rowSums(
                (X[-1, , drop = FALSE] - X[-nrow(X), , drop = FALSE])^2
            )
        )
    )
}

# Random rotation using safe and native ape infrastructure
random_rotate <- function(tree) {
    Ntip <- length(tree$tip.label)
    
    # Safe sampling of an internal node
    node <- sample((Ntip + 1):(Ntip + tree$Nnode), 1)
    
    # Properly rotates the node, preserving branch lengths and cladewise order
    ape::rotate(tree, node)
}

# Annealing
anneal_tree <- function(
    tree,
    traits,
    max_iter = 100000,
    temperature = 10,
    cooling = .099995, # Adjusted to match step counts smoothly
    patience = 5000
) {
    current <- tree
    current_cost <- trait_cost(current, traits)
    
    cat("Starting cost:", current_cost, "\n")
    
    best <- current
    best_cost <- current_cost
    no_change <- 0
    
    for (i in 1:max_iter) {
        candidate <- random_rotate(current)
        candidate_cost <- trait_cost(candidate, traits)
        delta <- candidate_cost - current_cost
        
        if (delta < 0 || runif(1) < exp(-delta / temperature)) {
            current <- candidate
            current_cost <- candidate_cost
        }
        
        if (current_cost < best_cost) {
            best <- current
            best_cost <- current_cost
            no_change <- 0
        } else {
            no_change <- no_change + 1
        }
        
        temperature <- temperature * cooling
        
        if (no_change > patience) {
            return(
                list(
                    tree = best,
                    cost = best_cost,
                    iterations = i
                )
            )
        }
    }
    
    list(
        tree = best,
        cost = best_cost,
        iterations = max_iter
    )
}


# Safely rebuilds the tree structure from scratch while preserving the visual layout
sanitize_tree <- function(tree) {
    ape::read.tree(text = ape::write.tree(tree))
}

############################################################
# METHOD 1:
# Random starting layout
############################################################

cat(
    "Initial cost:",
    trait_cost(my_tree,traits),
    "\n"
)

cat("\nRandom initialization\n")

random_order <- sample(my_tree$tip.label)

# sanitize_tree
random_tree <- sanitize_tree(rotateConstr(my_tree, random_order))

cat(
    "Initial random cost:",
    trait_cost(random_tree, traits),
    "\n"
)

start1 <- Sys.time()


random_result <- anneal_tree(
    random_tree,
    traits
)

time1 <- Sys.time()-start1



############################################################
# METHOD 2:
# PCA initialization
############################################################

cat("\nPCA initialization\n")


############################################################
# PCA scores
############################################################

pc <- prcomp(
    traits,
    scale.=TRUE
)


pc_order <- rownames(
    traits
)[
    order(pc$x[,1])
]


############################################################
# Rotate tree to PCA order
############################################################

# Added sanitize_tree()
pca_tree <- sanitize_tree(rotateConstr(my_tree, pc_order))





cat(
    "Initial PCA cost:",
    trait_cost(pca_tree,traits),
    "\n"
)



start2 <- Sys.time()

pca_result <- anneal_tree(
    pca_tree,
    traits
)

time2 <- Sys.time()-start2


############################################################
# METHOD 3:
# PCA + Greedy + Annealing
############################################################

cat("\nGreedy initialization\n")

# Greedy local optimization

greedy_optimize <- function(tree, traits, max_passes = 100){
    current <- tree
    current_cost <- trait_cost(current, traits)
    Ntip <- length(tree$tip.label)
    internal_nodes <- (Ntip + 1):(Ntip + tree$Nnode)
    
    pass <- 1
    
    repeat{
        improved <- FALSE
        
        for(node in internal_nodes){
            # Safely rotate using ape, maintaining cladewise integrity
            candidate <- ape::rotate(current, node)
            candidate_cost <- trait_cost(candidate, traits)
            
            if(candidate_cost < current_cost){
                current <- candidate
                current_cost <- candidate_cost
                improved <- TRUE
            }
        }
        
        cat("Greedy pass", pass, "cost:", current_cost, "\n")
        pass <- pass + 1
        
        if(!improved || pass > max_passes) break
    }
    
    list(tree = current, cost = current_cost, passes = pass - 1)
}

greedy_start <- greedy_optimize(my_tree, traits)
greedy_tree <- greedy_start$tree

cat(
    "Initial greedy cost:",
    greedy_start$cost,
    "\n"
)

start3 <- Sys.time()

greedy_result <- anneal_tree(
    greedy_tree,
    traits
)

time3 <- Sys.time()-start3

############################################################
# Compare
############################################################

results <- data.frame(

    Method=c(
        "Random start",
        "PCA + rotateConstr",
        "Greedy + Annealing"
    ),

    Initial_cost=c(
        trait_cost(random_tree, traits),
        trait_cost(pca_tree, traits),
        greedy_start$cost
    ),

    Final_cost=c(
        random_result$cost,
        pca_result$cost,
        greedy_result$cost
    ),

    Iterations=c(
        random_result$iterations,
        pca_result$iterations,
        greedy_result$iterations
    ),

    Time_seconds=c(
        as.numeric(time1),
        as.numeric(time2),
        as.numeric(time3)
    )

)

print(results)
```

The following code snippet contain the cost caculation function for evaluating the function that came before and after:

```{r}
get_layout_cost <- function(tree, traits) {
  n <- length(tree$tip.label)
  
  # 1. Vectorized Traversal (replaces recursive DFS)
  # Extracts tips in the exact order they appear in the edge matrix.
  # *CRITICAL*: Tree must be in "cladewise" order (see note below).
  is_tip <- tree$edge[, 2] <= n
  ord <- tree$tip.label[tree$edge[is_tip, 2]]
  
  # 2. Optimized Distance Calculation
  X <- traits[ord, , drop = FALSE]
  
  # diff() is a native C-level function that does exactly what X[-1,] - X[-nrow,] does, but faster.
  sum(sqrt(rowSums(diff(X)^2)))
}

get_layout_cost(greedy_result$tree,traits)
```


## Solution 6

The following code snippet finds the optimal tree according to the Bar-Joseph et al.
Using this algorithm can be a benchmark for the heuristic algorithm above:

```{r}
# =================================================================
# Optimal Leaf Ordering (OLO) for a phylogenetic tree, optimized
# against multi-dimensional continuous trait data
# =================================================================
#
# Standard OLO (Bar-Joseph et al. 2001, Bioinformatics) reorders the
# leaves of a binary tree -- by flipping the two children at each
# internal node -- to minimize the sum of dissimilarities between
# ADJACENT leaves in the final linear layout, WITHOUT changing the
# tree's topology (only left/right rotations are allowed).
#
# Here the topology is a fixed phylogeny (e.g. from sequence data),
# and the dissimilarity used to score adjacency comes from a
# separate multi-dimensional continuous trait matrix (morphology,
# gene expression, niche variables, etc.) -- NOT from patristic /
# cophenetic distances on the tree itself. The DP is agnostic to
# where the distance matrix comes from, so the classic algorithm
# applies directly; we just supply a trait-based distance matrix.
#
# Dependencies: only `ape`.
# Verified against brute-force enumeration of all 2^(n-1) rotations
# for n = 4..10 tips (exact match every time).
# =================================================================
 
library(ape)
 
# ---------------------------------------------------------------
# children[[node]] = integer vector of child node ids, for every
# node (tip or internal) in an ape "phylo" object.
# ---------------------------------------------------------------
get_children_list <- function(phy) {
  n_tip  <- length(phy$tip.label)
  n_node <- phy$Nnode
  children <- vector("list", n_tip + n_node)
  for (i in seq_len(nrow(phy$edge))) {
    p  <- phy$edge[i, 1]
    ch <- phy$edge[i, 2]
    children[[p]] <- c(children[[p]], ch)
  }
  children
}
 
# ---------------------------------------------------------------
# olo_phylo(phy, d)
#
#   phy : a fully bifurcating ape "phylo" object (n tips). Branch
#         lengths are irrelevant -- only topology matters.
#   d   : n x n distance matrix / dist object over the TRAITS,
#         with row/col i corresponding to phy$tip.label[i].
#
# Returns list(order, cost):
#   order = integer vector (permutation of 1:n) giving the leaf
#           order that minimizes summed adjacent trait-distance,
#           subject to the tree's topology.
#   cost  = the corresponding minimal total adjacent dissimilarity.
#
# Algorithm: Bar-Joseph dynamic programming. For every internal
# node v with children L, R, and for every pair of leaves (u,w)
# with u under L and w under R, define
#     M(u,w) = min_{b in leaves(L), c in leaves(R)}
#                 [ M_L(u,b) + M_R(c,w) + d(b,c) ]
# i.e. the cheapest way to lay out v's leaves starting at u and
# ending at w, where b/c are whichever leaves end up adjacent
# across the L/R seam. Because L's and R's leaf blocks can each be
# internally flipped independently, this exactly enumerates every
# topology-respecting leaf arrangement.
# ---------------------------------------------------------------
olo_phylo <- function(phy, d) {
 
  n_tip <- length(phy$tip.label)
  if (!is.matrix(d)) d <- as.matrix(d)
  stopifnot(nrow(d) == n_tip, ncol(d) == n_tip)
 
  children <- get_children_list(phy)
  memo <- vector("list", n_tip + phy$Nnode)
 
  dp <- function(node) {
    if (!is.null(memo[[node]])) return(memo[[node]])
    kids <- children[[node]]
 
    if (is.null(kids)) {                       # leaf / base case
      res <- list(leaves = node,
                   table = matrix(0, 1, 1, dimnames = list(node, node)))
      memo[[node]] <<- res
      return(res)
    }
    if (length(kids) != 2) {
      stop("Node ", node, " has ", length(kids), " children; the tree ",
           "must be fully bifurcating. Resolve polytomies first with ",
           "ape::multi2di().")
    }
 
    left  <- dp(kids[1])
    right <- dp(kids[2])
    Lleaves <- left$leaves;  Rleaves <- right$leaves
    LT <- left$table;        RT <- right$table
 
    # Re-embed each child's table as a full symmetric matrix over its
    # own leaves, so we can look up M(u,b) no matter which of the
    # child's own two sub-children u and b originally belonged to.
    L_full <- matrix(Inf, length(Lleaves), length(Lleaves),
                      dimnames = list(Lleaves, Lleaves))
    if (length(Lleaves) == 1) L_full[1,1] <- 0 else {
      L_full[rownames(LT), colnames(LT)] <- LT
      L_full[colnames(LT), rownames(LT)] <- t(LT)
    }
    R_full <- matrix(Inf, length(Rleaves), length(Rleaves),
                      dimnames = list(Rleaves, Rleaves))
    if (length(Rleaves) == 1) R_full[1,1] <- 0 else {
      R_full[rownames(RT), colnames(RT)] <- RT
      R_full[colnames(RT), rownames(RT)] <- t(RT)
    }
 
    tab  <- matrix(Inf, length(Lleaves), length(Rleaves),
                    dimnames = list(Lleaves, Rleaves))
    ptrB <- matrix(NA_integer_, length(Lleaves), length(Rleaves))
    ptrC <- matrix(NA_integer_, length(Lleaves), length(Rleaves))
 
    for (ui in seq_along(Lleaves)) {
      u <- Lleaves[ui]
      Lrow <- L_full[as.character(u), ]
      for (wi in seq_along(Rleaves)) {
        w <- Rleaves[wi]
        Rcol <- R_full[, as.character(w)]
        best <- Inf; bestB <- NA_integer_; bestC <- NA_integer_
        for (bi in seq_along(Lleaves)) {
          Lcost <- Lrow[bi]
          if (!is.finite(Lcost)) next
          b <- Lleaves[bi]
          total_vec <- Lcost + Rcol + d[as.character(b), as.character(Rleaves)]
          ci <- which.min(total_vec)
          if (total_vec[ci] < best) { best <- total_vec[ci]; bestB <- b; bestC <- Rleaves[ci] }
        }
        tab[ui, wi] <- best; ptrB[ui, wi] <- bestB; ptrC[ui, wi] <- bestC
      }
    }
 
    res <- list(leaves = c(Lleaves, Rleaves), table = tab,
                Lleaves = Lleaves, Rleaves = Rleaves,
                ptrB = ptrB, ptrC = ptrC, left = left, right = right)
    memo[[node]] <<- res
    res
  }
 
  root <- unique(setdiff(phy$edge[,1], phy$edge[,2]))
  stopifnot(length(root) == 1)
  top <- dp(root)
 
  # backtrack from the globally best (u,w) boundary pair at the root
  get_order <- function(node_res, start, end) {
    if (length(node_res$leaves) == 1) return(node_res$leaves)
    Lleaves <- node_res$Lleaves; Rleaves <- node_res$Rleaves
    if (start %in% Lleaves && end %in% Rleaves) {
      ui <- match(start, Lleaves); wi <- match(end, Rleaves)
      b <- node_res$ptrB[ui, wi]; c_ <- node_res$ptrC[ui, wi]
      c(get_order(node_res$left, start, b), get_order(node_res$right, c_, end))
    } else if (start %in% Rleaves && end %in% Lleaves) {
      rev(get_order(node_res, end, start))
    } else stop("invalid boundary pair")
  }
 
  min_cost <- min(top$table)
  idx <- which(top$table == min_cost, arr.ind = TRUE)[1, ]
  best_order <- get_order(top, top$Lleaves[idx[1]], top$Rleaves[idx[2]])
 
  list(order = as.integer(best_order), cost = min_cost)
}
 
# ---------------------------------------------------------------
# make_trait_dist(traits, phy): build a correctly ordered/labelled
# distance matrix from a trait table whose rownames are taxon names
# (in ANY row order) matching phy$tip.label.
# ---------------------------------------------------------------
make_trait_dist <- function(traits, phy, method = "euclidean", scale_traits = TRUE) {
  stopifnot(all(phy$tip.label %in% rownames(traits)))
  traits <- traits[phy$tip.label, , drop = FALSE]
  if (scale_traits) traits <- scale(traits)
  d <- as.matrix(dist(traits, method = method))
  rownames(d) <- colnames(d) <- as.character(seq_len(nrow(d)))
  d
}
 
# ---------------------------------------------------------------
# reorder_tree_olo(phy, d): run the DP and return a topology-
# identical phylo object whose tips are rotated into the OLO order
# (uses ape::rotateConstr, so no plotting package is required).
# ---------------------------------------------------------------
reorder_tree_olo <- function(phy, d) {
  res <- olo_phylo(phy, d)
  ord_labels <- phy$tip.label[res$order]
  phy_ord <- rotateConstr(phy, constraint = ord_labels)
  list(tree = phy_ord, order = ord_labels, cost = res$cost)
}

```

```{r}

d <- make_trait_dist(traits, my_tree, method = "euclidean", scale_traits = TRUE)
out <- reorder_tree_olo(my_tree, d)

cat("Minimal total adjacent trait-dissimilarity:", out$cost, "\n")
cat("Optimal tip order:\n"); print(out$order)

# out$tree is a normal phylo object -- plot it as usual:
plot(out$tree, main = "Tree with OLO-optimized tip order")

```


